---
title: "Robust lncRNA Discovery Workflow"
format: html
editor: visual
---

# Robust lncRNA Discovery Pipeline

This is an improved version of the lncRNA discovery workflow with enhanced robustness, error checking, and configuration management.

## Improvements Made

- **Configuration Management**: YAML-based configuration for different environments
- **Error Checking**: Comprehensive validation of files and processes
- **Restart Capability**: Ability to restart from any step
- **Logging**: Detailed logging with timestamps
- **Tool Detection**: Automatic detection of tools or manual configuration
- **Robust Downloads**: Retry logic for network operations
- **File Validation**: Check for corrupted or incomplete downloads
- **Progress Tracking**: Clear step-by-step progress reporting

```{r setup}
# Global R options
knitr::opts_chunk$set(echo = TRUE, error = TRUE)

# Load required libraries
suppressMessages({
  if (!require(yaml, quietly = TRUE)) install.packages("yaml")
  library(yaml)
})

# Configuration file path
CONFIG_FILE <- "config.yaml"

# Default configuration
default_config <- list(
  # General settings
  threads = 8,
  restart_from_step = 1,
  
  # Paths
  data_dir = "../data/12-Peve-lncRNA-discovery",
  output_dir = "../output/12-Peve-lncRNA-discovery", 
  log_dir = "../logs/12-Peve-lncRNA-discovery",
  
  # Data sources
  fastq_source = "https://gannet.fish.washington.edu/gitrepos/urol-e5/timeseries_molecular/E-Peve/output/01.00-E-Peve-RNAseq-trimming-fastp-FastQC-MultiQC/",
  fastq_suffix = "fq.gz",
  genome_source = "https://gannet.fish.washington.edu/seashell/snaps/Porites_evermanni_v1.fa",
  gtf_source = "https://github.com/urol-e5/timeseries_molecular/raw/99f0563a067ca9d010cb206dfd44b36d8f77de00/E-Peve/data/Porites_evermanni_validated.gtf",
  gff_source = "https://github.com/urol-e5/timeseries_molecular/raw/99f0563a067ca9d010cb206dfd44b36d8f77de00/E-Peve/data/Porites_evermanni_validated.gff3",
  
  # Analysis parameters
  gff_pattern = 'class_code "u"|class_code "x"|class_code "o"|class_code "i"',
  min_transcript_length = 200,
  min_coverage = 10,
  min_sample_fraction = 0.5,
  
  # Tool paths
  tool_paths = list(
    auto_detect = TRUE,
    raven = list(
      hisat2_dir = "/home/shared/hisat2-2.2.1/",
      samtools_dir = "/home/shared/samtools-1.12/",
      stringtie_dir = "/home/shared/stringtie-2.2.1.Linux_x86_64/",
      gffcompare_dir = "/home/shared/gffcompare-0.12.6.Linux_x86_64/",
      bedtools_dir = "/home/shared/bedtools2/bin/",
      cpc2_dir = "/home/shared/CPC2_standalone-1.0.1/bin/",
      conda_path = "/opt/anaconda/anaconda3/bin/conda",
      featurecounts_path = "/home/shared/subread-2.0.5-Linux-x86_64/bin/featureCounts"
    ),
    klone = list(
      hisat2_dir = "",
      samtools_dir = "",
      stringtie_dir = "",
      gffcompare_dir = "/srlab/programs/gffcompare-0.12.6.Linux_x86_64/",
      bedtools_dir = "",
      cpc2_dir = "/srlab/programs/CPC2_standalone-1.0.1/bin/",
      conda_path = "/mmfs1/gscratch/srlab/nextflow/bin/miniforge/bin/conda",
      featurecounts_path = ""
    )
  )
)

# Load or create configuration
if (file.exists(CONFIG_FILE)) {
  config <- read_yaml(CONFIG_FILE)
  cat("Configuration loaded from", CONFIG_FILE, "\n")
} else {
  cat("Creating default configuration file...\n")
  write_yaml(default_config, CONFIG_FILE)
  cat("Default configuration written to", CONFIG_FILE, "\n")
  config <- default_config
}

# Merge with defaults for any missing values
config <- modifyList(default_config, config)

# Create directories
for (dir in c(config$data_dir, config$output_dir, config$log_dir)) {
  if (!dir.exists(dir)) {
    dir.create(dir, recursive = TRUE)
    cat("Created directory:", dir, "\n")
  }
}
```

```{r helper-functions}
# Helper functions for robust workflow execution

# Logging function
log_message <- function(message, level = "INFO") {
  timestamp <- format(Sys.time(), "%Y-%m-%d %H:%M:%S")
  log_entry <- sprintf("[%s] %s: %s", timestamp, level, message)
  cat(log_entry, "\n")
  
  # Write to log file
  log_file <- file.path(config$log_dir, "workflow.log")
  cat(log_entry, "\n", file = log_file, append = TRUE)
}

# Step tracking
current_step <- 1
step_names <- c(
  "Setup and dependency check",
  "Data download",
  "File validation",
  "HISAT2 alignment",
  "StringTie assembly",
  "GFFCompare analysis",
  "Candidate filtering",
  "CPC2 coding potential",
  "Final processing",
  "Count matrix generation"
)

should_run_step <- function(step_number) {
  return(step_number >= config$restart_from_step)
}

start_step <- function(step_number, step_name) {
  if (!should_run_step(step_number)) {
    log_message(sprintf("Skipping Step %d: %s", step_number, step_name), "SKIP")
    return(FALSE)
  }
  
  log_message(sprintf("Starting Step %d: %s", step_number, step_name), "START")
  current_step <<- step_number
  return(TRUE)
}

complete_step <- function(step_number, step_name) {
  log_message(sprintf("Completed Step %d: %s", step_number, step_name), "COMPLETE")
}

# File validation functions
check_file_exists <- function(filepath, description = "") {
  if (!file.exists(filepath)) {
    log_message(sprintf("Missing file: %s (%s)", filepath, description), "ERROR")
    return(FALSE)
  }
  return(TRUE)
}

check_file_not_empty <- function(filepath, description = "") {
  if (!check_file_exists(filepath, description)) return(FALSE)
  
  size <- file.info(filepath)$size
  if (is.na(size) || size == 0) {
    log_message(sprintf("Empty file: %s (%s)", filepath, description), "ERROR")
    return(FALSE)
  }
  return(TRUE)
}

check_html_download <- function(filepath, description = "") {
  if (!check_file_exists(filepath, description)) return(FALSE)
  
  # Check first few lines for HTML content
  first_lines <- readLines(filepath, n = 3, warn = FALSE)
  if (any(grepl("<html>|<HTML>|<!DOCTYPE", first_lines, ignore.case = TRUE))) {
    log_message(sprintf("HTML file detected instead of expected format: %s (%s)", filepath, description), "ERROR")
    return(FALSE)
  }
  return(TRUE)
}

# Tool detection and validation
detect_tool <- function(tool_name, possible_paths = c()) {
  # First try system PATH
  system_path <- Sys.which(tool_name)
  if (system_path != "") {
    log_message(sprintf("Found %s in system PATH: %s", tool_name, system_path))
    return(dirname(system_path))
  }
  
  # Try possible paths
  for (path in possible_paths) {
    full_path <- file.path(path, tool_name)
    if (file.exists(full_path) && file.access(full_path, mode = 1) == 0) {
      log_message(sprintf("Found %s at: %s", tool_name, path))
      return(path)
    }
  }
  
  log_message(sprintf("Tool not found: %s", tool_name), "WARNING")
  return("")
}

# Environment detection
detect_environment <- function() {
  hostname <- Sys.info()["nodename"]
  
  if (grepl("raven", hostname, ignore.case = TRUE)) {
    return("raven")
  } else if (grepl("klone", hostname, ignore.case = TRUE)) {
    return("klone")
  } else {
    log_message("Unknown environment, using default tool paths", "WARNING")
    return("raven")  # Default to raven paths
  }
}
```

```{r setup-environment}
# Step 1: Setup and dependency check
if (start_step(1, step_names[1])) {
  
  # Create directories
  dirs_to_create <- c(
    config$data_dir,
    config$output_dir,
    config$log_dir,
    file.path(config$data_dir, "fastq")
  )
  
  for (dir in dirs_to_create) {
    if (!dir.exists(dir)) {
      dir.create(dir, recursive = TRUE)
      log_message(sprintf("Created directory: %s", dir))
    }
  }
  
  # Set up tool paths
  if (config$tool_paths$auto_detect) {
    env <- detect_environment()
    log_message(sprintf("Detected environment: %s", env))
    
    if (env %in% names(config$tool_paths)) {
      tool_config <- config$tool_paths[[env]]
    } else {
      tool_config <- config$tool_paths$raven
    }
  } else {
    # Use manual configuration
    env <- detect_environment()
    tool_config <- config$tool_paths[[env]]
  }
  
  # Validate critical tools
  critical_tools <- c("hisat2", "samtools", "stringtie", "bedtools", "python")
  missing_tools <- c()
  
  for (tool in critical_tools) {
    tool_path <- detect_tool(tool, c(
      tool_config[[paste0(tool, "_dir")]],
      "/usr/bin", "/usr/local/bin", "/opt/bin"
    ))
    if (tool_path == "") {
      missing_tools <- c(missing_tools, tool)
    }
  }
  
  if (length(missing_tools) > 0) {
    log_message(sprintf("Missing critical tools: %s", paste(missing_tools, collapse = ", ")), "ERROR")
    log_message("Please install missing tools or update configuration", "ERROR")
    stop("Critical tools missing")
  }
  
  # Set environment variables
  env_vars <- list(
    THREADS = as.character(config$threads),
    DATA_DIR = config$data_dir,
    OUTPUT_DIR = config$output_dir,
    LOG_DIR = config$log_dir,
    FASTQ_DIR = file.path(config$data_dir, "fastq"),
    FASTQ_SOURCE = config$fastq_source,
    FASTQ_SUFFIX = config$fastq_suffix,
    GENOME_SOURCE = config$genome_source,
    GTF_SOURCE = config$gtf_source,
    GFF_SOURCE = config$gff_source,
    GFFPATTERN = config$gff_pattern,
    GENOME_FASTA = file.path(config$data_dir, "genome.fasta"),
    GENOME_GTF = file.path(config$data_dir, "genome.gtf"),
    GENOME_GFF = file.path(config$data_dir, "genome.gff"),
    GENOME_INDEX = file.path(config$output_dir, "genome.index"),
    MIN_TRANSCRIPT_LENGTH = as.character(config$min_transcript_length),
    MIN_COVERAGE = as.character(config$min_coverage),
    MIN_SAMPLE_FRACTION = as.character(config$min_sample_fraction)
  )
  
  # Add tool paths
  for (name in names(tool_config)) {
    env_vars[[toupper(name)]] <- tool_config[[name]]
  }
  
  # Set environment variables
  do.call(Sys.setenv, env_vars)
  
  log_message("Environment setup completed")
  complete_step(1, step_names[1])
}
```

For brevity, I'll continue the improved workflow with the remaining critical improvements. Let me complete the file with the remaining steps:

<function_calls>
<invoke name="str_replace_editor">
<parameter name="command">str_replace